Shapes analysis for time series.

### 2024-05-22 NEURIPS24/paper.tex
update sam

% One option to tackle this issue is to derive ...
% feature representation of time series which depends on the problem at hand.
% which is parti... -> which is of prime interest in ...

\section{Introduction}
%https://arxiv.org/abs/1811.00075
%Appliqué sur les "UWaveGestureLibrary","ArticularyWordRecognition","Cricket" pour l'archive UEA
%"ArrowHead", "BME", "ECG200","FacesUCR","GunPoint" dans le cas univarié

@@

    %  \end{itemize}
    Our contributions can be summarized as follows:
    \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset. In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters, also providing guidelines for tuning.
       \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset.
        \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets.
 \end{itemize}

@@

\section{Related Works}
Shape analysis focuses on statistical analysis of mathematical objects invariant under some deformations like rotations, dilations, or time parameterization.
   The main idea is to represent these objects in a complete Riemannian manifold $(\mathcal{M},\mathbf{g})$ with a metric $\mathbf{g}$ adapted to the geometry of the problem \cite{miller2006geodesic}.
    Then, any set of points in $\mathcal{M}$ can be represented as points in the tangent space of their Frechet mean $\mathbf{m}_0$ \cite{pal2017riemannian,le2001locating} by considering their logarithms.
     The goal is to find a well-suited Riemannian structure according to the nature of the studied object.

       Leaving the LDDMM representation, the results of \cite{srivastava2010shape,heo2024logistic} addresses the representation of curves with the Square-Root Velocity (SRV) representation.
        However, the SRV representation is applied after a reparametrization of the temporal dimension on the unit length segment.
         Consequently, the graph structure of the time series is not respected, and the original time evolution of the time series is not encoded in the final representation.
          Very recently, in a functional data analysis framework, a paper \cite{wu2024shape} (Shape-FPCA) improved by representing the original time evolution.
           Nevertheless, this method is made for \textit{continuous objects} and only applies to time series of \textit{same length}, making the estimation more sensitive to noise.
  %However, the case of time series has not been adequately addressed, considering the special attention the time axis requires.
  %For instance, the SRV representation applies after a reparametrization in time, such that the original time evolution of the time series is not represented in the final features.
% Depending on the tasks, e.g classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep} or clustering , a meaningful vector representation of time series can be derived.
%  Initially, in this research, handcrafted features were computed from time series, such as the mean, wavelets \cite{percival2000wavelet}, or Fourier coefficients \cite{bloomfield2004fourier}, depending on expert knowledge.
%   However, this approach is time-consuming and potentially biased by priors.
%   Observing the success of neural networks, features are automatically extracted by choosing an objective to optimize and a neural architecture tailored to the problem.
%    Taking a set of time series as input, features can be identified by solving specific supervised tasks—classification \cite{ismail2019deep}, anomaly detection \cite{blazquez2021review}, forecasting \cite{torres2021deep}— or in an unsupervised manner by deriving labels from the inputs \cite{meng2023unsupervised}.
%    \thi{Rajouter une image, et une phrase par rapport au temps}
%    Although the deep learning approach is particularly effective for supervised tasks, annotated datasets are rare, making unsupervised methods more practical.
%    These methods have proven effective in identifying different patterns in the data and providing a vector representation that segregates well into clusters \cite{meng2023unsupervised}.
%     However, they may lack precision on small and homogeneous datasets with only one typical shape, often arising from clinical experiments addressing a specific phenomenon (refer to figures).
%      While current unsupervised methods separate representations into clusters, our focus is on the inter-individual variability within one cluster.
%       Typically, we aim to perform Principal Components Analysis (PCA) \cite{bro2014principal} on individual patterns with similar shapes. The representation is dependant on the input dataset.

 %\citet{srivastava2010shape,heo2024logistic} address the representation of curves $C:\msi \to \Rset^d$, $\msi\subset \Rset$, having unitary velocity by using the Square-Root Velocity (SRV) representation.
 %However, the case of time series has not been adequately addressed, considering the special attention the time axis requires.
 %For instance, the SRV representation applies after a reparametrization in time, such that the original time evolution of the time series is not represented in the final features.
 %In this regard, this work is not an application of LDDMM to 1D curves \cite{glaunes2008large}.
 %Compared to Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} or Continuous Normalizing Flows \cite{chen2018neural,grathwohl2019scalable,salman2018deep} for diffeomorphisms learning, the number of hyperparameters to tune is minimal, and the related optimization problem is well-posed.
 Balancing between discrete and continuous elements is a challenging task.
  In the deep learning literature \cite{chen2018neural, kidger2020neural, tzen2019neural, jia2019neural, liu2019neural, ansari2023neural}, Neural Ordinary Differential Equations (Neural ODEs) \cite{chen2018neural} learn continuous latent representations using a vector field parameterized by a neural network, serving as a continuous analogue to Residual Networks \cite{zagoruyko2016wide}.
   This approach was further enhanced by Neural Controlled Differential Equations (Neural CDEs) \cite{kidger2020neural} for handling irregular time series, functioning as continuous-time analogs of RNNs \cite{schuster1997bidirectional}.
 Extending Neural ODEs, Neural Stochastic Differential Equations (Neural SDEs) introduce regularization effects \cite{liu2019neural}, although optimization remains challenging.
  Leveraging techniques from continuous-discrete filtering theory, Ansari et al. \cite{ansari2023neural} applied succesfully Neural SDEs to irregular time series.
  Oh \textit{et al.} \cite{oh2024stable} improved these results by incorporating the concept of controlled paths into the drift term, similar to how Neural CDEs outperform Neural ODEs.
  With TS-LDDMM, the representation is also derived from an ODE, but the velocity field is parameterized with kernels and optimized to have a minimal norm, which enhance interpretability.

\section{Notations}
We denote by integer ranges by $[k:l]=\{k,\ldots,l\}\subset \mathcal{P}(\Zset)$ and $ [l]=[1:l]$ with $k,l\in \Nset$,
by $\rmC^m(\msi,\mse)$ the set of $m$-times continously differentiable function defined on an open set $\msu$ to a normed vector space $\mse$,
 by $||u||_\infty=\inf_{x\in \msu} |u(x)| $ for any bounded function $u:\msu \to \mse$,
and by $\Nset_{>0}$ is the set of positive integers. 
%We are in a sub class of curve quotiented by temporal reparametrisation since a time series' graph can be understood as the equivalent class of t ->(t,s(t))

%Reproducing Kernel Hilbert Space (RKHS)
\section{Background on LDDMM}
\label{section:LDDMM}

@@

% LDDMM is designed to analyzed an existing dataset, while NF and CNF are made to generalize a dataset for data augmentation.
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{equation}
  \label{eq:LDDMM_dynamic}
    \frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,
    \phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{equation}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $
and $\msv$ is a Hilbert space of continuously differentiable function
on $\Rset^{d'}$.  If
$||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any
$u\in \msv$ and
$v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||
v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport}
$\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
 Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. 
This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

@@

\paragraph{Discrete parametrization of diffeomorpshim.}
%\alain{adapt this without mentioning applicaiton to time seriess}
%\sam{Est ce qu'on devrait pas introduire la dicretization du graphe ici et dire que l'on peut utiliser le théoreme de représentation}
In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). % $K(x,y)\triangleq \exp(-|x-y|^2/\sigma^2)\operatorname{Id}_{d'}$ for any $(x,y)\in(\Rset^{d'})^2$, where $\sigma^2>0$ is the kernel width and $\operatorname{Id}_{d'}$ is the identity matrix.
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \citet{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

@@

%  \end{algorithmic}
%  \end{algorithm}
\paragraph{Relation to Continuous Normalizing Flows.}

One particular popular choice to address the problem of learning a diffeomorphism or a velocity field is Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} (NF) or their continuous counterpart \cite{chen2018neural,grathwohl2019scalable,salman2018deep} (CNF).
However, we do not rely on this class of learning algorithms for several reasons. Indeed, existing and simple normalizing flows are not suitable for the type of data that we are interested in this paper \cite{feng2023multi,deng2020modeling}. 
  In addition,  they are primarily designed to have tractable Jacobian functions, while we do not require such property in our applications. 
Finally, the use of a differential flow solution of an ODE
\eqref{eq:LDDMM_dynamic} trick is also at the basis of CNF, which
then consists of learning a velocity field to address in fitting the data through a loss aiming to address the problem
at hand. Nevertheless, the main difference between CNF and LDDMM lies in the
parametrization of the velocity field. LDDMM uses kernels to
derive closed form formula and enhance interpretability while
NF and CNF take advantage of deep neural networks to scale with
large dataset in high dimensions.
\section{Methodology}
%\paragraph{Assumptions}
%Denoting by $C^0(\msi,\msj) $ the space of continous functions between two sets $\msi,\msj$,

We consider in this paper observations which consist in a population of $N$ multivariate time series, for any $j\in[N]$, $s^j \in \rmC^1(\msi_j,\Rset^{d})$. 
However, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$.
 Note that \textbf{the number of samples $n_j$ is not necessary the same accross individuals} and the timestamps can be \textbf{irregularly sampled}.
%To ease the presentation of our methodology, we suppose that we have access to the continuous time series $(s^j)_{j\in[N]}$, while in practice, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$.
 %However, this difference does not change the overall rationale of our method and its implementation up to minor changes, which we summarize in \eqref{eq:general_optimization_problem}.
 We assume the time series population is globally homogeneous regarding their "shapes" even if inter-individual variability exists.
 Intuitively speaking, the "shape" of a time series $s:\msi\to \Rset^d$ is encoded in its graphs $\msg(s)$ defined as the set $\{(t,s(t)):\eqsp t\in\msi \} $ and not only in its values $s(\msi)=\{s(t):\eqsp t\in\msi \} $ since the time axis is crucial. % on the shape as illustrated in Figure\alain{ (mettre ref)}.
 As a motivating use-case, $s^j$ can be the time series of a heartbeat extracted from an individual's electrocardiogram (ECG), see \Cref{fig:transport}.
 The homogeneity in a resulting dataset comes from the fact that humans have similar shapes of heartbeat \cite{ye2012heartbeat,madona2021pqrst}.
 % the shape is also referred as morphology of the time series. 
  %  [More specifically, the heterogeneity in such dataset is often due to local or glocal dilatation of the time series regarding time, the heartbeat is quicker or slower depending on the individual.
  %  In maths, denoting by $\mcd(\msi,\msj) $ the set of diffeomorpshim defined on the sets $\msi$ to $\msj$,
  %   there exist a common pattern $\mathbf{s}_0: \msi \to \Rset^d$ and individual time reparametrisations $ \psi_j\in\mcd(\msi,\msi_j) $ such that,
  %   for any $j\in[N]$, $s^j=\mathbf{s}_0\circ \psi_j^{-1} $.]
%  \paragraph{Goals}%\sam{potentiellement changer l'ordre de time and space}
\paragraph*{The deformation problem.}
In this paper, we aim to study the inter-individual variability in the dataset by finding a relevant representation of each time series.
Inspired from the framework of shape analysis \cite{vaillant2004statistics}, addressing similar problems in morphology,
 we suggest to represent each time series' graph $\msg(s^j)$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$, related to a time series $\mathbf{s}_0:\msi \to\Rset^d$, by a diffeomorphism $\phi_j$ on $\Rset^{d+1}$, for any $j\in[N]$,
\begin{equation}
 \label{eq:transformation}
 \phi_j.\msg(\mathbf{s}_0)=\{\phi_j\left(t,\mathbf{s}_0(t)\right), \eqsp t\in \msi \} \eqsp.
\end{equation}
$\bfs_0$ will be understood as the typical representative shape common to the collection of time series $(s^j)_{j\in[N]}$.
As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s^j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$.
We aim to learn $\msg(\bfs_0)$ and $(\phi_j)_{j\in[N]} $.

@@

Nevertheless, \eqref{eq:general_optimization_problem} ask to define a kernel and a loss in order to perform geodesic shooting \ref{eq:relaxation}, which is the purpose of the next subsection.
   \subsection{Application of LDDMM to time series analysis: TS-LDDMM}
%\tcr{Mettre l'applicaiton clair pour les times series}

@@

Similarly, selecting a loss function on sets $\mathscr{L}$ that considers the temporal evolution in a time series' graph is crucial for meaningful comparisons with time series data. Consequently, we introduce the oriented Varifold distance.

       \paragraph{A representation separating space and time.}
       We prove that two time series graphs can always be linked by a time transformation composed of a space transformation. Moreover, a time series graph transformed by this kind of transformation is always a time series graph.
        We define $\Psi_\gamma\in \mcd(\Rset^{d+1}) : (t,x)\in\Rset^{d+1}\to (\gamma(t),x)$ for any $\gamma\in \mcd(\Rset)$ and $\Phi_f:  (t,x)\in\Rset^{d+1}\to (t,f(t,x)) $ for any $f\in \rmC^1(\Rset^{d+1},\Rset^d)$. 
        We have the following representation theorem.
        All proofs are given in \Cref{appendix:proofs}.

@@

%\sam{améliorer transition}
\paragraph{Loss}

@@

\sam{Parler de méthode adaptatif ici}

@@

\include{experiment}

>>>>

% One option to tackle this issue is to derive ...
% feature representation of time series which depends on the problem at hand.
% which is parti... -> which is of prime interest in ...
\vspace{-1ex}
\section{Introduction}
\vspace{-1ex}
%https://arxiv.org/abs/1811.00075
%Appliqué sur les "UWaveGestureLibrary","ArticularyWordRecognition","Cricket" pour l'archive UEA
%"ArrowHead", "BME", "ECG200","FacesUCR","GunPoint" dans le cas univarié

@@

    %  \end{itemize}
    Our contributions can be summarized as follows:
    \begin{itemize}
       \item We propose an unsupervised method (TS-LDDMM) to analyze the inter-individual variability of shapes in a time series dataset (\Cref{section:methodology}). In particular, the method can handle multivariate time series \textit{irregularly sampled} and with \textit{variable sizes}.
       \item We motivate our extension of LDDMM to time series by introducing a theoretical framework with a representation theorem for time series graph (\Cref{theorem:representation}) and kernels related to their structure (\Cref{lemma:choice_of_kernel_V}).
       \item We demonstrate the identifiability of the model by estimating the true generating parameter of synthetic data, and we highlight the sensitivity of our method concerning its hyperparameters (\Cref{appendix:identifiability}), also providing guidelines for tuning (\Cref{appendix:kernel_TS_LDDMM}).
       \item We highlight the \textit{interpretability} of TS-LDDMM for studying the inter-individual variability in a clinical dataset (\Cref{section:experiments}).
        \item We illustrate the quantitative interest of such representation on classification tasks on real shape-based datasets with regular and irregular sampling (\Cref{appendix:classification_shape_analysis}).
 \end{itemize}

@@

  \vspace{-1ex}
\section{Notations}
We denote by integer ranges by $[k:l]=\{k,\ldots,l\}\subset \mathcal{P}(\Zset)$ and $ [l]=[1:l]$ with $k,l\in \Nset$,
by $\rmC^m(\msi,\mse)$ the set of $m$-times continously differentiable function defined on an open set $\msu$ to a normed vector space $\mse$,
 by $||u||_\infty=\inf_{x\in \msu} |u(x)| $ for any bounded function $u:\msu \to \mse$,
and by $\Nset_{>0}$ is the set of positive integers. 
%We are in a sub class of curve quotiented by temporal reparametrisation since a time series' graph can be understood as the equivalent class of t ->(t,s(t))

%Reproducing Kernel Hilbert Space (RKHS)
\vspace{-1ex}
\section{Background on LDDMM}
\vspace{-1ex}
\label{section:LDDMM}

@@

 \vspace{-1ex}
% LDDMM is designed to analyzed an existing dataset, while NF and CNF are made to generalize a dataset for data augmentation.
\paragraph{Representing diffeomorpshims as deformations.}
Such \textit{deformations} $\phi$ are constructed via differential flow equations, for any $x_0\in \Rset^{d'} $ and $\tau\in[0,1]$:
\begin{equation}
  \label{eq:LDDMM_dynamic}
    \frac{\dd X(\tau)}{\dd \tau}= v_\tau(X(\tau)), \quad X(0)=x_0\eqsp ,
    \phi^v_\tau(x_0)=X(\tau), \quad \phi^v=\phi^v_1  \eqsp ,
\end{equation}
where the velocity field is $v:\tau\in [0,1]\mapsto v_\tau\in \msv $
and $\msv$ is a Hilbert space of continuously differentiable function
on $\Rset^{d'}$.  If
$||\dd u ||_{\infty}+|| u ||_{\infty}\leq ||u ||_\msv $ for any
$u\in \msv$ and
$v\in \rml^2([0,1],\msv)=\{v\in \rmC^0([0,1],\msv): \int_0^1 ||
v_\tau||^2_\msv \dd \tau<\infty \} $, by \citep[Theorem 5]{glaunes2005transport}
$\phi^v$ exists and belongs to $\mcd(\Rset^{d'})$, where we denote by $\mcd(\mso) $ the set of diffeomorpshim defined on an open set $\mso$ to $\mso$.
 Therefore, for any choice of $v$, $\phi^v$ defines a valid deformation. 
This offers a general recipe to construct diffeomorphism given a functional space $\msv$.

@@

\vspace{-1ex}
\paragraph{Discrete parametrization of diffeomorpshim.}
%\alain{adapt this without mentioning applicaiton to time seriess}
%\sam{Est ce qu'on devrait pas introduire la dicretization du graphe ici et dire que l'on peut utiliser le théoreme de représentation}
In this part, $\msv$ is chosen as an RKHS \cite{berlinet2011reproducing} generated by a smooth kernel $K$ (e.g., Gaussian). % $K(x,y)\triangleq \exp(-|x-y|^2/\sigma^2)\operatorname{Id}_{d'}$ for any $(x,y)\in(\Rset^{d'})^2$, where $\sigma^2>0$ is the kernel width and $\operatorname{Id}_{d'}$ is the identity matrix.
We follow \cite{durrleman2013sparse} and define a 
 discrete parameterization of the velocity fields to perform geodesics shooting \eqref{eq:geodesics_shooting}.  
  The initial velocity field $v_0$ is chosen as a finite linear combination of the RKHS basis vector fields, 
$\mathbf{n}_0$ control points $\msx_0=(x_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0}$ and momentum vectors $\alpha_0=(\alpha_{k,0})_{k\in[\mathbf{n}_0]}\in (\Rset^{d'})^{\mathbf{n}_0} $ are defined such that for any $x\in \Rset^{d'}$, 
  \begin{equation}
    \label{eq:def_v0}
    v_0\left(\alpha_0,\msx_0\right)(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_{k,0})\alpha_{k,0} \eqsp .
  \end{equation}
   In our applications, the control points $(x_{k,0})_{k\in[\mathbf{n}_0]}$ can be understood as the discretized graph $(t_k,\mathbf{s}_0(t_k))_{k\in[\mathbf{n}_0]}$ of a starting time series $\mathbf{s}_0$. 
  With this parametrization of $v_0$, \citet{miller2006geodesic} show that the velocity field $v$ of the solution of \eqref{eq:geodesics_shooting} keeps the same
  structure along time, such that for any $x\in\Rset^{d'}$ and $\tau\in[0,1]$, 
  \begin{equation}
    \label{eq:specific_form}
    v_\tau(x)=\sum_{k=1}^{\mathbf{n}_0} K(x,x_k(\tau))\alpha_{k}(\tau) \eqsp ,
  \end{equation}
  \begin{equation} 
    \label{eq:integration}
      \left\{
        \begin{aligned}
        & \frac{\dd x_k(\tau)}{\dd \tau}=v_\tau(x_k(\tau)) \eqsp, \quad
        \frac{\dd \alpha_k(\tau)}{\dd \tau}=-\sum_{k=1}^{\mathbf{n}_0} \dd_{x_k(\tau)} K(x_k(\tau),x_l(\tau))\alpha_{l}(\tau)^\top \alpha_{k}(\tau) \eqsp  \\
        & \alpha_k(0)=\alpha_{k,0},\quad x_k(0)=x_{k,0} \eqsp , k\in[\mathbf{n}_0] 
        \end{aligned}
        \right .
  \end{equation}
  These equations are derived from the hamiltonian $H:(\alpha_k,x_k)_{k\in [\mathbf{n}_0]}\mapsto \sum_{k,l=1}^{\mathbf{n}_0} \alpha_{k}^\top K(x_k,x_l)\alpha_{l}  $, such that
  the velocity norm is preserved $||v_\tau||_{\msv}=||v_0||_\msv $ for any $\tau\in [0,1]$.
   By \eqref{eq:integration}, the velocity field related to a geodesic $v^*$ is fully parametrized by its initial control points and momentum $(x_{k,0},\alpha_{k,0})_{k\in[\mathbf{n}_0]}$.
   Thus, given a set of targets $\mathbf{y}=(y_i)_{i\in[T_2]}$ in $\Rset^{d'}$, a set of starting points $\mathbf{x}=(x_{i,0})_{i\in[T_1]}$ in $\Rset^{d'}$, a RKHS's kernel $K:\Rset^{d'}\times \Rset^{d'}\to \Rset^{d'\times d'}$, a distance on sets $\mathscr{L}$, 
 a numerical integration scheme of ODE and a penalty factor $\lambda>0$, the basic geodesic shooting step minimizes the following function using a gradient descent method:
   \begin{equation}
    \label{eq:relaxation}
    \mathcal{F}_{\mathbf{x},\mathbf{y}}: (\alpha_k)_{k\in [T_1]}\mapsto \mathscr{L}\left(\varphi^{\{v_0\}}.\mathbf{x},\mathbf{y}\right)+\lambda||v_0||_\msv^2 \eqsp,  
   \end{equation}
   where $v_0$ is defined by \eqref{eq:def_v0} and $\varphi^{\{v_0\}}.\mathbf{x}$ is the result of the numerical integration of \eqref{eq:integration} using control points $\mathbf{x}$ and initial momentums $(\alpha_k)_{k\in[T_1]} $.

@@

%  \end{algorithmic}
%  \end{algorithm}
\vspace{-1ex}
\paragraph{Relation to Continuous Normalizing Flows.}

One particular popular choice to address the problem of learning a diffeomorphism or a velocity field is Normalizing Flows \cite{rezende2015variational,kobyzev2020normalizing} (NF) or their continuous counterpart \cite{chen2018neural,grathwohl2019scalable,salman2018deep} (CNF).
However, we do not rely on this class of learning algorithms for several reasons. Indeed, existing and simple normalizing flows are not suitable for the type of data that we are interested in this paper \cite{feng2023multi,deng2020modeling}. 
  In addition,  they are primarily designed to have tractable Jacobian functions, while we do not require such property in our applications. 
Finally, the use of a differential flow solution of an ODE
\eqref{eq:LDDMM_dynamic} trick is also at the basis of CNF, which
then consists of learning a velocity field to address in fitting the data through a loss aiming to address the problem
at hand. Nevertheless, the main difference between CNF and LDDMM lies in the
parametrization of the velocity field. LDDMM uses kernels to
derive closed form formula and enhance interpretability while
NF and CNF take advantage of deep neural networks to scale with
large dataset in high dimensions.
\vspace{-1ex}
\section{Methodology}
\label{section:methodology}
%\paragraph{Assumptions}
%Denoting by $C^0(\msi,\msj) $ the space of continous functions between two sets $\msi,\msj$,
\vspace{-1ex}
We consider in this paper observations which consist in a population of $N$ multivariate time series, for any $j\in[N]$, $s^j \in \rmC^1(\msi_j,\Rset^{d})$. 
However, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$.
 Note that \textbf{the number of samples $n_j$ is not necessary the same accross individuals} and the timestamps can be \textbf{irregularly sampled}.
%To ease the presentation of our methodology, we suppose that we have access to the continuous time series $(s^j)_{j\in[N]}$, while in practice, we can only access a $n_j$-samples $\tsig^j=(\tsig_i^j=s^j(t^j_i))_{i\in[n_j]}$ collected at timestamps $(t^j_i)_{i\in[n_j]}$ for any $j \in [N]$.
 %However, this difference does not change the overall rationale of our method and its implementation up to minor changes, which we summarize in \eqref{eq:general_optimization_problem}.
 We assume the time series population is globally homogeneous regarding their "shapes" even if inter-individual variability exists.
 Intuitively speaking, the "shape" of a time series $s:\msi\to \Rset^d$ is encoded in its graphs $\msg(s)$ defined as the set $\{(t,s(t)):\eqsp t\in\msi \} $ and not only in its values $s(\msi)=\{s(t):\eqsp t\in\msi \} $ since the time axis is crucial. % on the shape as illustrated in Figure\alain{ (mettre ref)}.
 As a motivating use-case, $s^j$ can be the time series of a heartbeat extracted from an individual's electrocardiogram (ECG), see \Cref{fig:transport}.
 The homogeneity in a resulting dataset comes from the fact that humans have similar shapes of heartbeat \cite{ye2012heartbeat,madona2021pqrst}.
 % the shape is also referred as morphology of the time series. 
  %  [More specifically, the heterogeneity in such dataset is often due to local or glocal dilatation of the time series regarding time, the heartbeat is quicker or slower depending on the individual.
  %  In maths, denoting by $\mcd(\msi,\msj) $ the set of diffeomorpshim defined on the sets $\msi$ to $\msj$,
  %   there exist a common pattern $\mathbf{s}_0: \msi \to \Rset^d$ and individual time reparametrisations $ \psi_j\in\mcd(\msi,\msi_j) $ such that,
  %   for any $j\in[N]$, $s^j=\mathbf{s}_0\circ \psi_j^{-1} $.]
%  \paragraph{Goals}%\sam{potentiellement changer l'ordre de time and space}
\vspace{-1ex}
\paragraph*{The deformation problem.}
In this paper, we aim to study the inter-individual variability in the dataset by finding a relevant representation of each time series.
Inspired from the framework of shape analysis \cite{vaillant2004statistics}, addressing similar problems in morphology,
 we suggest to represent each time series' graph $\msg(s^j)$ as the transformation of a reference graph $\msg(\mathbf{s}_0)$, related to a time series $\mathbf{s}_0:\msi \to\Rset^d$, by a diffeomorphism $\phi_j$ on $\Rset^{d+1}$, for any $j\in[N]$,
\begin{equation}
 \label{eq:transformation}
 \phi_j.\msg(\mathbf{s}_0)=\{\phi_j\left(t,\mathbf{s}_0(t)\right), \eqsp t\in \msi \} \eqsp.
\end{equation}
$\bfs_0$ will be understood as the typical representative shape common to the collection of time series $(s^j)_{j\in[N]}$.
As $\bfs_0$ is supposed to be fixed, then the representation of the time series $(s^j)_{j\in[N]}$ boils down to the one of the transformation $(\phi_j)_{j\in[N]}$.
We aim to learn $\msg(\bfs_0)$ and $(\phi_j)_{j\in[N]} $.

@@

Nevertheless, \eqref{eq:general_optimization_problem} ask to define a kernel and a loss in order to perform geodesic shooting \ref{eq:relaxation}, which is the purpose of the next subsection.
\vspace{-1ex}
   \subsection{Application of LDDMM to time series analysis: TS-LDDMM}
   \vspace{-1ex}
%\tcr{Mettre l'applicaiton clair pour les times series}

@@

Similarly, selecting a loss function on sets $\mathscr{L}$ that considers the temporal evolution in a time series' graph is crucial for meaningful comparisons with time series data. Consequently, we introduce the oriented Varifold distance. 
\vspace{-1ex}
       \paragraph{A representation separating space and time.}
       We prove that two time series graphs can always be linked by a time transformation composed of a space transformation. Moreover, a time series graph transformed by this kind of transformation is always a time series graph.
        We define $\Psi_\gamma\in \mcd(\Rset^{d+1}) : (t,x)\in\Rset^{d+1}\to (\gamma(t),x)$ for any $\gamma\in \mcd(\Rset)$ and $\Phi_f:  (t,x)\in\Rset^{d+1}\to (t,f(t,x)) $ for any $f\in \rmC^1(\Rset^{d+1},\Rset^d)$. 
        We have the following representation theorem.
        All proofs are given in \Cref{appendix:proofs}.

@@

%\sam{améliorer transition}
\vspace{-1ex}
\paragraph{Loss}

@@

%\sam{Parler de méthode adaptatif ici}

@@

\vspace{-1ex}
\input{experiment}
\vspace{-1ex}

\section{Related Works}
\vspace{-1ex}
Shape analysis focuses on statistical analysis of mathematical objects invariant under some deformations like rotations, dilations, or time parameterization.
   The main idea is to represent these objects in a complete Riemannian manifold $(\mathcal{M},\mathbf{g})$ with a metric $\mathbf{g}$ adapted to the geometry of the problem \cite{miller2006geodesic}.
    Then, any set of points in $\mathcal{M}$ can be represented as points in the tangent space of their Frechet mean $\mathbf{m}_0$ \cite{pal2017riemannian,le2001locating} by considering their logarithms.
     The goal is to find a well-suited Riemannian structure according to the nature of the studied object.
